---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clock

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/DavisVaughan/clock/branch/master/graph/badge.svg)](https://codecov.io/gh/DavisVaughan/clock?branch=master)
[![R-CMD-check](https://github.com/DavisVaughan/clock/workflows/R-CMD-check/badge.svg)](https://github.com/DavisVaughan/clock/actions)
<!-- badges: end -->

The goal of clock is to provide a comprehensive library for working with dates and times. A few of the key features of clock are:

- _Time points_ with precision that can vary from second up to nanosecond.

- _Calendar_ types, such as year-month-day, year-month-weekday, year-quarter-day, and iso-year-week-day, that can vary in precision from year up to nanosecond.

- _Duration_ types for working with absolute intervals of time.

- Clear and unambiguous handling of otherwise confusing date time manipulation issues, such as leap days and daylight savings.

- A high level API supporting R's Date and POSIXct classes to ease you into clock before having to learn about the more advanced features mentioned above.

## Installation

You can install the development version of clock with:

``` r
# install.packages("remotes")
remotes::install_github("DavisVaughan/clock")
```

## Example

```{r, warning=FALSE, message=FALSE}
library(clock)
library(magrittr)
```

With clock, there is a high level API supporting Date and POSIXct. This is often the easiest way to begin using clock.

For example, to add 5 days:

```{r}
x <- as.Date("2019-05-01")

add_days(x, 5)
```

To add 5 hours:

```{r}
y <- as.POSIXct("2019-01-01 02:30:00", tz = "America/New_York")

add_hours(y, 5)
```

To get the month:

```{r}
get_month(x)

get_month(y)
```

To set the day:

```{r}
set_day(x, 1:5)

# Last day in the month
set_day(x, "last")
```
