---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clock

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/DavisVaughan/clock/branch/master/graph/badge.svg)](https://codecov.io/gh/DavisVaughan/clock?branch=master)
[![R-CMD-check](https://github.com/DavisVaughan/clock/workflows/R-CMD-check/badge.svg)](https://github.com/DavisVaughan/clock/actions)
<!-- badges: end -->

clock is a new R package for working with date-time data. It:

- Provides a new family of orthogonal date-time classes (durations, time points, zoned times, and calendars) that partition responsibilities so that you only have to think about time zones when you need them.

- Implements a high level API for Date and POSIXct classes that lets you get productive quickly without having to learn the details of the underlying classes.

- Requires explicit handling of impossible dates (e.g. what date is one month after January 31st?) and ambiguous times (caused by daylight savings).

- Is built on the C++ [date](https://github.com/HowardHinnant/date) library, which provides a correct and high-performance backend.

There are four key classes in clock, inspired by the design of the C++ date and chrono libraries. Some types are more efficient than others at particular operations, and it is only when all 4 are taken as a whole do you get a complete date time library.

- A _duration_ counts units of time, like “5 years” or “6 nanoseconds”. Bigger units are defined in terms of seconds, i.e. 1 day is 86400 seconds and 1 year is 365.2425 days. Durations are important because they form the backbone of clock; it’s relatively rare to use them directly.

- A _time point_ records an instant in time, like “1:24pm January 1st 2015”. It combines a _duration_ with a “clock” that defines when to start counting and what exactly to count. There are two important types of time in clock: `sys_time` and `naive_time`. They’re equivalent until you start working with zoned times.

- A _zoned time_ is a time point paired with a time zone. You can create them from either a `sys_time` or a `naive_time`, depending on whether you want to convert the time point from UTC (leaving the underlying duration unchanged, but changing the printed time), or declare that the time point is in a specific time zone (leaving the printed time unchanged, but changing the underlying duration). Zoned times are primarily needed for communication with humans.

- A _calendar_ represents a date using a combination of fields like year-month-day, year-month-weekday, year-quarter-day, or iso-year-week-day, along with hour/minute/second fields to represent time within a day (so they’re similar to R’s POSIXlt). Calendar objects are extremely efficient at arithmetic involving irregular periods such as months, quarters, and years and at getting and setting specified components. A calendar can represent invalid dates (like 2020-02-31) which only need to be resolved when converting back to a time point.

## Installation

You can install the development version of clock with:

``` r
# install.packages("remotes")
remotes::install_github("DavisVaughan/clock")
```

## Example

```{r, warning=FALSE, message=FALSE}
library(clock)
library(magrittr)
```

With clock, there is a high level API supporting Date and POSIXct. This is often the easiest way to begin using clock.

For example, to add 5 days:

```{r}
x <- as.Date("2019-05-01")

add_days(x, 5)
```

To add 5 hours:

```{r}
y <- as.POSIXct("2019-01-01 02:30:00", tz = "America/New_York")

add_hours(y, 5)
```

To get the month:

```{r}
get_month(x)

get_month(y)
```

To set the day:

```{r}
set_day(x, 1:5)

# Last day in the month
set_day(x, "last")
```

## Acknowledgements

The ideas used to build clock have come from a number of places:

- First and foremost, clock depends on and is inspired by the [date](https://github.com/HowardHinnant/date) library by Howard Hinnant, a variant of which has been voted in to C++20.

- The R "record" types that clock is built on come from [vctrs](https://github.com/r-lib/vctrs).

- The [nanotime](https://github.com/eddelbuettel/nanotime) package was the first to implement a nanosecond resolution time type for R.

- The [zoo](https://cran.r-project.org/web/packages/zoo/index.html) package was the first to implement year-month and year-quarter types, and has functioned as a very successful time series infrastructure package for many years.
